---
title: "Using Computational Tools Besides R"
subtitle: "How I Learned to Use SQL and Love GitHub"
author: "Mauricio 'Pachá' Vargas · PUC Chile"
institute: "University of Toronto Students Union's Workshops"
date: "2021-03-16"
output:
  xaringan::moon_reader:
    includes:
      in_header: "mathjax-equation-numbers.html"
    lib_dir: libs
    css: ["xaringan-themer.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: ""
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(emojifont) 
library(xaringanthemer)
style_mono_accent(
  base_color = "#002a5c",
  header_font_google = google_font("Fira Sans"),
  text_font_google = google_font("Fira Sans"),
  code_font_google = google_font("Fira Mono"),
  text_font_size = "1.25rem",
  colors = c(white = "#FFFFFF", gray = "#898C9C"),
  extra_css = list(
    ".large" = list(
      "font-size" = "2.75rem"
    ),
    ".inverse > h1 > code" = list(
      "color" = "var(--inverse-header-color)"
    ),
    ".colophon" = list(
      "font-size" = "0.9rem",
      "position" = "absolute",
      "bottom" = 0
    ),
    ".flex-box" = list(
    "display" = "flex",
    "flex-direction" = "column",
    "justify-content" = "space-between",
    "height" = "70%"
    ),
    # the first p of .flext-box should have no top-margin
    ".flex-box > p:first-child" = list(
      "margin-top" = 0
    ),
    ".flex-box > p:last-of-type" = list(
      "padding-bottom" = "1em"
    ),
    ".seventyfive" = list(
      "width" = "75%",
      "margin" = "0 auto"
    ),
    ".title-slide h1" = list(
      "font-size" = "4rem",
      "margin-bottom" = "0rem"
      ),
    ".title-slide h2" = list(
      "font-size" = "1.75rem",
      "margin-top" = "0rem",
      "margin-bottom" = "4rem"
      ),
    ".title-slide h3" = list(
      "font-size" = "1.75rem",
      "margin-block-start" = "0rem",
      "margin-block-end" = "0rem"
      ),
    ".title-slide > h3:last-of-type" = list(
      "font-size" = "1rem",
      "margin-block-start" = "4rem",
      "margin-block-end" = "2rem"
    )
  )
)
```

---

# Required packages

Check that you have installed the following packages:

* `library(usethis)` ✔
* `library(HistData)` ✔
* `library(RPostgres)` ✔
* `library(askpass)` ✔
* `library(ggplot2)` ✔
* `library(pkgdown)` ✔

---

background-position: 50% 50%
background-size: cover
class: center, middle

# Now run create_project("01-models-review")

---

# Linear models review

Let $\renewcommand{\vec}[1]{\boldsymbol{#1}} \newcommand{\R}{\mathbb{R}} \vec{y} \in \R^n$ be the outcome and $X \in \mathbb{R}^{n\times p}$ be the design matrix in the context of a linear model with intercept:

$$\vec{y} = X\vec{\beta} + \vec{e}$$

Being: 

$$\underset{n\times 1}{\vec{y}} = \begin{pmatrix}y_0 \cr y_1 \cr \vdots \cr y_n\end{pmatrix} \text{ and }
\underset{n\times p}{X} = \begin{pmatrix}1 & x_{11} & & x_{1p} \cr 1 & x_{21} & & x_{2p} \cr & \ddots & \cr 1 & x_{n1} & & x_{np}\end{pmatrix} = (\vec{1} \: \vec{x}_1 \: \ldots \: \vec{x}_p)$$

---

# Linear models review

In linear models the aim is to minimize the error term by choosing $\hat{\vec{\beta}}$. One possibility is to minimize the squared error by solving this optimization problem:

$$
\begin{equation}
\label{min}
\displaystyle \min_{\vec{\beta}} S = \|\vec{y} - X\vec{\beta}\|^2
\end{equation}
$$

Books such as [Baltagi](https://www.amazon.com/Econometrics-Springer-Texts-Business-Economics/dp/3642200583) discuss how to solve $\eqref{min}$ and other equivalent approaches that result in this optimal estimator:

$$
\begin{equation}
\label{beta}
\hat{\vec{\beta}} = (X^tX)^{-1} X^t\vec{y}
\end{equation}
$$
---

# Linear models review

With one independent variable and intercept, this is $y_i = \beta_0 + \beta_1 x_{i1} + e_i$, equation $\eqref{beta}$ means:
$$
\begin{equation}
\label{beta2}
\hat{\beta}_1 = cor(\vec{y},\vec{x}) \cdot \frac{sd(\vec{y})}{sd(\vec{x})} \text{ and } \hat{\beta}_0 = \bar{y} - \hat{\beta}_1 \bar{\vec{x}}
\end{equation}
$$
---

# Linear models review

Consider the model:
$$mpg_i = \beta_0 + \beta_1 wt_i + \beta_2 cyl_i + e_i$$

Now create the script `01-linear-models.R` and fit the model:

```{r anova_mtcars_example1, eval = FALSE}
lm(mpg ~ wt + cyl, data = mtcars)
```

---

# Linear models review

Create a function that returns the next matrix for any $(x_1, x_2)$ such as:

```{r, eval = FALSE}
x1 <- mtcars$wt
x2 <- mtcars$cyl

my_design_matrix <- function(x1, x2) {
  x0 <- rep(1, length(x1))
  X <- cbind(x0,x1,x2)
  return(X)
}
```

---

# Linear models review

It's the same to use `lm` or to perform a matrix multiplication because of equation $\eqref{beta}$. Let's use the previous function to compare:

```{r, eval = FALSE}
y <- mtcars$mpg

fit <- lm(y ~ x1 + x2)
coefficients(fit)

X <- my_design_matrix(x1,x2)
beta <- solve(t(X) %*% X) %*% (t(X) %*% y)
beta
```

---

# Linear models review

Verify $\eqref{beta2}$ by using R commands and the Galton dataset:

```{r, eval = FALSE}
library(HistData)

y <- Galton$child
x <- Galton$parent
beta1 <- cor(y, x) *  sd(y) / sd(x)
beta0 <- mean(y) - beta1 * mean(x)

c(beta0, beta1)

lm(y ~ x)
```

---

# Linear models review

Consider the model:

$$y_i - \bar{y} = \beta_0 + \beta_1 (x_{i1} - \bar{x}_{1}) + \beta_2 (x_{i2} - \bar{x}_{2}) + e_i$$

A particular case of this would be:

```{r, eval = FALSE}
new_y <- mtcars$mpg - mean(mtcars$mpg)
new_x1 <- mtcars$wt - mean(mtcars$wt)
new_x2 <- mtcars$cyl - mean(mtcars$cyl)

fit2 <- lm(new_y ~ new_x1 + new_x2)
summary(fit2)
```

What can we say about $\beta_0$?

---

# Linear models review

Now fitting the model:
$$y_i - \bar{y}_i = \beta_1 (x_{i1} - \bar{x}_{i1}) + \beta_2 (x_{i2} - \bar{x}_{i2}) + e_i$$
```{r, eval = FALSE}
fit3 <- lm(new_y ~ new_x1 + new_x2 + 0)
# or fit3 <- lm(new_y ~ new_x1 + new_x2 - 1)
summary(fit3)
```

---

background-position: 50% 50%
background-size: cover
class: center, middle

# 2 minutes break

---

background-position: 50% 50%
background-size: cover
class: center, middle

# Now run create_project("02-sql-connections")

---

# Using SQL from R

* Explore https://databases.pacha.dev/#intendo-database 
* Load usethis, dplyr and RPostgres
* Create the environmental variables `dbedu_usr` and `dbedu_pwd` with `edit_r_environ()`

---

# Using SQL from R

* Create an R script `con1` with `use_r("con1")`
* Create a connection object in the script that uses the environmental variables

```{r, eval = FALSE}
con1 <- dbConnect(
  Postgres(),
  user = Sys.getenv("dbedu_usr"),
  password = Sys.getenv("dbedu_pwd"),
  dbname = "intendo",
  host = "databases.pacha.dev"
)
```

---

# Using SQL from R

Use any of the connections to obtain a subset from the `users` table containing

* 🇨🇦 as `country`
* Users of the `android` `platform`
* And that were `acquired` from `organic` growth

```{r, eval = FALSE}
canada_android_organic <- tbl(con1, "users") %>% 
  filter(
    country == "Canada",
    platform == "android",
    acquired == "organic"
  ) %>% 
  collect()
```

---

# Using SQL from R

Free exercise: Explore the different countries, platforms and acquisition sources. Then compare your query with the previous one and comment freely.

Just a tiny help:
```{r, eval = FALSE}
tbl(con1, "users") %>% 
  select(platform) %>% 
  distinct()
```

---

background-position: 50% 50%
background-size: cover
class: center, middle

# 5 minutes break

---

background-position: 50% 50%
background-size: cover
class: center, middle

# Now run create_package("mypackage")
# Participation shall be very important
